---
title: "Getting started with CEAS"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with CEAS}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
bibliography: refs.bib
link-citations: yes
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

## Setup

```{r setup}
library(ceas)
```

## Importing Seahorse rates data

The `read_data` function takes a list of Excel files. An easy way to get
such a list is to put all your data in a directory and list its contents. Here
we use the package's internal datasets, but `list.files` will take a directory
name as its first argument.

```{r data}
rep_list <- system.file("extdata", package = "ceas") |>
  list.files(pattern = "*.xlsx", full.names = TRUE)
raw_data <- readxl::read_excel(rep_list[1], sheet = 2)
knitr::kable(head(raw_data))
```

The data requires the following columns: `r colnames(raw_data)`. The `Group`
column needs to be in the format `biological_group<space>Assay_type` as shown
above. Upon reading with `read_data`, the `Group` column is split into two
`group` and `assay` columns on the space. This output format can be set in the
Seahorse machine before starting the experiment. If you already have the data,
this column will have to be converted to this format to work with ceas.

```{r read_dataformat}
seahorse_rates <- read_data(rep_list)
knitr::kable(head(seahorse_rates))
```

### Normalization

There are two types of normalization involved in Seahorse data analysis. One is
background normalization done by the Wave software. *ceas* will produce a
warning if it finds that the "Background" data are not 0 (see first row of the
table above).

The other is biological normalization based on the cell count or the mass of
protein. If the data is not already biologically normalized, you will need a csv
file containing experimental groups and cell counts or $\mu$g of protein in this
format:

```{r, norm_csv}
norm_csv <- system.file("extdata", package = "ceas") |>
  list.files(pattern = "norm.csv", full.names = TRUE)
read.csv(norm_csv) |> knitr::kable()
```

Your csv file's full path may be passed into `read_data()` using the `norm`
argument.

```{r normalized_read}
read_data(rep_list, norm = norm_csv) |> head() |> knitr::kable()
```

## Calculating energetics

### Partitioning data

The energetics calculation workflow involves partitioning the data into its time
point and assay intervals.

```{r partition_data}
partitioned_data <- partition_data(seahorse_rates)
```

#### Alternative data formats {.tabset}

While the default options are set for an experiment with both a mitochondrial
and glycolysis assay, if you have only a mitochondrial assay or no glycolysis
assay, the `assay_types` list parameter can be modified to account for that.

##### Mito + Glyco (our default)

```{r, eval = FALSE}
partitioned_data <- partition_data(
  seahorse_rates,
  assay_types = list(
    basal = "MITO",
    uncoupled = "MITO",
    maxresp = "MITO",
    nonmito = "MITO",
    no_glucose_glyc = "GLYCO",
    glucose_glyc = "GLYCO",
    max_glyc = "GLYCO"
  ),
  basal_tp = 3,
  uncoupled_tp = 6,
  maxresp_tp = 8,
  nonmito_tp = 12,
  no_glucose_glyc_tp = 3,
  glucose_glyc_tp = 6,
  max_glyc_tp = 8
)
```

##### Data in the form of @mookerjee2017:

```{r, eval = FALSE}
partitioned_data <- partition_data(
  seahorse_rates,
  assay_types = list(
    basal = "MITO",
    uncoupled = "MITO",
    maxresp = NA,
    nonmito = "MITO",
    no_glucose_glyc = NA,
    glucose_glyc = "MITO",
    max_glyc = "MITO"
  ),
  basal_tp = 5,
  uncoupled_tp = 10,
  maxresp_tp = NA,
  nonmito_tp = 12,
  no_glucose_glyc_tp = 2,
  glucose_glyc_tp = 5,
  max_glyc_tp = 10
)
```

##### Just Mito

```{r, eval = FALSE}
partitioned_data <- partition_data(
  seahorse_rates,
  assay_types = list(
    basal = "MITO",
    uncoupled = "MITO",
    maxresp = "MITO",
    nonmito = "MITO",
    no_glucose_glyc = NA,
    glucose_glyc = "MITO",
    max_glyc = "MITO"
  ),
  basal_tp = 3,
  uncoupled_tp = 6,
  maxresp_tp = 8,
  nonmito_tp = 12,
  no_glucose_glyc_tp = NA,
  glucose_glyc_tp = 3,
  max_glyc_tp = 6
)
```

####

Note that the time point parameters (`maxresp_tp` and `no_glucose_glyc_tp`) also
need to be changed accordingly.

The `get_energetics` function requires pH, pK$_a$ and buffer values.

```{r get_energetics}
energetics <- get_energetics(partitioned_data, ph = 7.4, pka = 6.093, buffer = 0.10)
```

For more information on the calculations see the article on [ATP
calculations](atp_calculation.html).

## Plotting

### Bioenergetic scope plot

The `bioscope_plot` function plots a 2D representation of group “bioenergetic
scope.” Bioenergetic scope describes the theoretical energetic space in which a
matrix operates. The bioenergetic scope coordinates are JATP from OXPHOS on the
y-axis and JATP from glycolysis on the x-axis. The points represent mean basal
and/or max JATP from OXPHOS and glycolysis and the vertical and horizontal lines
represent the standard deviation or confidence interval of JATP from OXPHOS or
glycolysis, respectively.

```{r bioscope_plot}
(bioscope <- bioscope_plot(energetics))
```

### Rate plots {.tabset}

The `rate_plot` function provides an overview of OCR or ECAR over time, which
enables cross-group energetic comparisons before and after the addition of
energetic-modulating compounds. The `rate_plot` line represents mean group OCR
or ECAR over the sequential measurements (x-axis) and the shaded variance region
represents standard deviation or specified confidence interval.

#### Oxygen consumption rate (OCR)

```{r ocr}
(ocr <- rate_plot(seahorse_rates, measure = "OCR"))
```

#### Extracellular Acidification Rate (ECAR)

```{r ecar}
(ecar <- rate_plot(seahorse_rates, measure = "ECAR"))
```

### ATP plots {.tabset}

The `atp_plot` function plots group JATP values, which enables cross-group
OXPHOS and glycolytic JATP comparisons at basal and max conditions. The
`atp_plot` symbols represent the mean basal or max JATP from OXPHOS or
glycolysis, and the crossbar boundaries represent the standard deviation or
confidence interval JATP variance.

#### Basal glycolysis

```{r basal_glyc}
(basal_glyc <- atp_plot(energetics, basal_vs_max = "basal", glyc_vs_resp = "glyc"))
```

#### Basal respiration

```{r basal_resp}
(basal_resp <- atp_plot(energetics, basal_vs_max = "basal", glyc_vs_resp = "resp"))
```


#### Max glycolysis

```{r max_glyc}
(max_glyc <- atp_plot(energetics, basal_vs_max = "max", glyc_vs_resp = "glyc"))
```


#### Max respiration

```{r max_resp}
(max_resp <- atp_plot(energetics, basal_vs_max = "max", glyc_vs_resp = "resp"))
```

### Customizing plots

CEAS is designed to work with existing `ggplot2` customization functionality and
doesn't include more than shape and size options for its plots.

For example, to change the colors used in the plot, simply make the plot and
add the custom colors you'd like:

#### Colors

```{r custom_colors}
custom_colors <- c("#e36500", "#b52356", "#3cb62d", "#328fe1")
```

```{r}
bioscope +
ggplot2::scale_color_manual(
  values = custom_colors
)
```

```{r}
ocr +
ggplot2::scale_color_manual(
  values = custom_colors
)
```

#### Labels {.tabset}

##### Change axis labels

```{r}
ecar +
    ggplot2::labs(x = "Time points")
```

##### Change label size

```{r}
basal_glyc +
    ggplot2::theme(axis.text = ggplot2::element_text(size = 20))
```

#### Editing functions

We are working on making the plots as customizable as possible. However, if
there are options that cannot be set in the calls to the plotting functions or
with `ggplot2` functions, you can get the code used to make the plots by running
the function name without parenthesis and modify it. Further, since every step
in the ceas workflow provides a dataset, you can run the modified function or
your own custom plotting functions on those datasets.

```{r, eval = FALSE}
rate_plot
```


```{r, results = 'asis', echo = FALSE}
func_code <- capture.output(dput(rate_plot))
cat("```r\n")
cat(func_code, sep = "\n")
cat("\n```")
```

In RStudio, you can run `utils::edit` to modify a function.

```{r, eval = FALSE}
edit(rate_plot)
```



## References
